An Algebraic Theory File Synchronization

### 2015-12-25 2015/axioms.tex
Improved definitions

So that we could describe the effects of commands in general,
let us introduce some notation
and note some observations.
We already know that
commands usually do not occur in isolation,
and are applied to filesystems in time.
Therefore we will investigate sequences of commands with a well-defined order.
\begin{mydef}[$\cc$ and $\emptyseq$]
We use ``$\cc$'' to concatenate commands to form a sequence, or concatenate sequences to form a longer sequence:
$S=\cbdaa{n}{\vald}\cc\cbfaa{m}{\valf}$, or $S'=S\cc\cbba{o}$,
with the meaning that the command on the left is applied first:
\[ \alpha\cc\beta\aFS = \beta(\alpha\aFS). \]
Sequences of commands form a free semigroup, and, as usual,
$\emptyseq$ denotes its unit element, the empty sequence,
which by definition leaves all filesystems unchanged.
\end{mydef}

>>>>

So that we could describe the effects of commands in general,
let us introduce some notation
and note some observations.
We already know that
commands usually do not occur in isolation,
and are applied to filesystems in time.
Therefore we will investigate sequences of commands with a well-defined order.
\begin{mydef}[Sequences of commands: $\cc$ and $\emptyseq$]
We use ``$\cc$'' to concatenate commands to form a sequence, or concatenate sequences to form a longer sequence:
$S=\cbdaa{n}{\vald}\cc\cbfaa{m}{\valf}$, or $S'=S\cc\cbba{o}$,
with the meaning that the command on the left is applied first:
\[ \alpha\cc\beta\aFS = \beta(\alpha\aFS). \]
Sequences of commands form a free semigroup, and, as usual,
$\emptyseq$ denotes its unit element, the empty sequence,
which by definition leaves all filesystems unchanged.
\end{mydef}

### 2015-12-25 2015/def-commands.tex
Improved definitions

Next we define commands on filesystems.
As described above, we aimed to select a set of commands
that captures as much information as possible
about the operations, and is also symmetric.
We start by summarizing our reasons for doing so.

@@

We therefore have four pieces of information related to each command:
its input type, its output type, the node to which it is applied, and finally
the value that it stores at the node.
We will denote commands using quadruples that list these
in a the following format:
% TODO convert into definition
\[ \cxynv, \]
where $X$ is the input type, $Y$ is the output type, $n$ is the node,
and $\valvy$ is a value of the appropriate type from $\setvx{Y}$.
We use $\cchard$, $\ccharf$ or $\ccharb$ to represent the different
types. For example, $\cbdaa{n}{\vald}$ 
will represent $mkdir(n,\vald)$.
% TODO Just say "omit"
Alternatively, we use a triple of the two types and a node if 
there is only a single suitable output value for the given type (e.g. $\empt$).
That is, $\cdba{n}$ will represent $rmdir(n)$.

@@

\noindent
A command can only succeed if the original value at node $n$ has a type that matches
the input type of the command. If this is not the case, or if the resulting
filesystem no longer has the tree property, then we say that the command
{\em breaks} the filesystem. Broken filesystems are considered equal
(but not equal to any working filesystem), and we note them by $\FS=\fsbroken$.

So that we could reason about sequences of commands that break every filesystem, 
we introduce the command $\cbrk$ that simply breaks every filesystem.

@@

% TODO include command in def
\begin{mydef}[Replacement operator]
Using $\FS[\valv/n]$ we denote a filesystem derived from $\FS$:
\[ \FS[\valv/n](m) :=
   \begin{cases}
   \valv &\mbox{if~} m=n\\
   \FS(m) &\mbox{otherwise.}
   \end{cases}
\]
\end{mydef}

The effect of the commands is as follows:
\begin{align*}
&\cbrk\aFS = \fsbroken \\
&\cxynv\aFS = 
   \begin{cases}
   \fsbroken &\mbox{if~} \FS=\fsbroken\\
   \fsbroken &\mbox{if~} \FS(n)\not\in\setvx{X}\\
   \fsbroken &\mbox{if~} \FS[\valvy/n] \mbox{~violates the tree property}\\
   \FS[\valvy/n] &\mbox{otherwise.}
   \end{cases}
\end{align*}

>>>>

Next we define commands on filesystems.
As described above, we aimed to select a set of commands
that captures as much information
about the operations as possible, and is also symmetric.
We start by summarizing our reasons for doing so.

@@

\begin{mydef}[Regular commands]
Therefore we have four pieces of information related to each command,
and represent them using quadruples that list
their input type ($X$), output type ($Y$),
the node to which they are applied ($n$),
and the value they store at the node ($\valvy$):
\[ \cxynv, \]
where $\valvy\in\setvx{Y}$.

We use $\cchard$, $\ccharf$ or $\ccharb$ to represent the different
types, and omit the final value if it is determined by the output type
(when it is $\empt$ or $\vald$).
For example, $\cbdaa{n}{\vald}$ will represent $mkdir(n,\vald)$,
and $\cdba{n}$ will represent $rmdir(n)$.
\end{mydef}

@@

\noindent
A command can only succeed if the original value at node $n$ has a type that matches
the input type of the command. If this is not the case, or if the resulting
filesystem no longer has the tree property, then we say that the command
{\em breaks} the filesystem, and assigns $\fsbroken$ to it.
Broken filesystems are considered equal, but not equal to any working filesystem.

So that we could reason about sequences of commands that break every filesystem, 
we introduce, in addition to the regular commands,
the command $\cbrk$ that simply breaks every filesystem.

@@

\begin{mydef}[Effect of commands]
We use a \emph{replacement operator} of the form
$\FS[\valv/n]$ to denote a filesystem derived from $\FS$:
\[ \FS[\valv/n](m) :=
   \begin{cases}
   \valv &\mbox{if~} m=n\\
   \FS(m) &\mbox{otherwise.}
   \end{cases}
\]
The effect of the commands is then as follows:
\begin{align*}
&\cbrk\aFS = \fsbroken \\
&\cxynv\aFS = 
   \begin{cases}
   \fsbroken &\mbox{if~} \FS=\fsbroken\\
   \fsbroken &\mbox{if~} \FS(n)\not\in\setvx{X}\\
   \fsbroken &\mbox{if~} \FS[\valvy/n] \mbox{~violates the tree property}\\
   \FS[\valvy/n] &\mbox{otherwise.}
   \end{cases}
\end{align*}
\end{mydef}

### 2015-12-25 2015/def-filesystem.tex
Reordering; improvements; new notation; improved figure
Improved definitions

We model a filesystem using a function $\FS$ with a set of nodes (filesystem paths) $\setn$ as its domain,
and a set of possible contents or values, $\setv$, as its codomain:
\begin{mydef}[Filesystem]
\[ \FS : \setn \rightarrow \setv \] 
\end{mydef}
In our model, $\setn$ 
serves as a namespace or ``skeleton'' for the filesystem, and it
contains all possible nodes, including the ones where there is no file or directory.
To model empty nodes, $\setv$ 
contains a special element ($\empt$) which is the value of $\FS$ at these nodes.
We consider the contents of files, as well as any meta-information of files
and directories (e.g. permissions and other flags) part of the values in $\setv$.

The nodes form a disjoint union of rooted directed trees.
Tao et al. in \cite{TSR} describe a similar filesystem model, although
they also model inodes and restrict the filesystem to a just single tree.
\begin{mydef}[$\parent$]
The function $\parent(n)$ returns the parent node of $n$, or
returns $\topnode$ if $n$ is the root of a tree.
\end{mydef}

Every filesystem has a so-called \emph{tree property}, which means that
if the filesystem is not empty at a node, and the node has a parent,
then there must be a directory at the parent node.

To model this, in $\setv$ we distinguish between 
file values ($\setf$) and directory values ($\setd$), that is,
if $\setb = \{\empt\}$ then:
\[ \setv = \setb \cup \setf \cup \setd \]
The tree property can then be defined as
\begin{mydef}[Tree property]
\begin{align*}
\forall n\in\setn: &\FS(n) \neq \empt \\ % TODO alignment problem
&\quad\wedge \parent(n) \neq \topnode \\
&\Rightarrow \FS(\parent(n)) \in \setd. 
\end{align*}
\end{mydef}
This means that as we move down from the root of a tree of nodes,
the types of values we encounter in the filesystem can only change according to the
transition diagram in \cref{fig_transition}, from directories ($\cchard$) to files ($\ccharf$)
to empty nodes ($\ccharb$).

@@

In this paper, $n$, $m$ and $o$ denote nodes in $\setn$.
We use $\valf$ for an arbitrary element in $\setf$, 
and $\vald$ for an arbitrary element in $\setd$,
and by $\valvx$ we mean a value of type $X$ in $\setvx{X}$.

% Simplification
% --------------

A simplification we will assume in this paper is regarding directories.
As Bill Zissimopoulos pointed 
out,\footnote{Bill Zissimopoulos, personal communication, August 2015.} 
we often do not want to consider metadata stored in
directories (e.g. permission settings) during synchronization,
as these are not generally understood well by users,
and, if needed, conflict resolution on these settings can be easily automated.
In other words, the contents of directories are all equal,
which can be modelled by assuming that $|\setd|=1$.

>>>>

We model a working filesystem
using a function $\FS$ with a set of nodes (filesystem paths) $\setn$ as its domain,
and a set of possible contents or values $\setv$ as its codomain.
Alternatively, if the filesystem has encountered a problem and is no longer usable, 
we assign it the special value $\fsbroken$:
\begin{mydef}[Filesystem]
\begin{gather*}
\FS
\begin{cases}
\setn \rightarrow \setv \\
\fsbroken
\end{cases}
\end{gather*}
\end{mydef}
In our model, $\setn$ 
serves as a namespace or ``skeleton'' for the filesystem, and it
contains all possible nodes, including the ones where there is no file or directory,
which we denote using a special value $\empt\in\setv$.

The nodes form a disjoint union of rooted directed trees,
which determines the following $\parent$ function.
Tao et al. in \cite{TSR} describe a similar filesystem model, although
they also model inodes and restrict the filesystem to a just single tree.
\begin{mydef}[$\parent$]
The function $\parent(n)$ returns the parent node of $n$, or
returns $\topnode$ if $n$ is the root of a tree.
\end{mydef}

Every working filesystem has a so-called \emph{tree property}, which means that
if the filesystem is not empty at a node, and the node has a parent,
then there must be a directory at the parent node.
To model this, in $\setv$ we select another special value
to represent directories, as we assume that apart from their location
in the filesystem, all directories are equal.
We do this because, as Bill Zissimopoulos pointed 
out (\cite{BZ}),
we often do not want to consider metadata stored in
directories (e.g. permission settings) during synchronization,
as these are not generally understood well by users,
and, if needed, conflict resolution on these settings can be easily automated.
We will also see that this assumption makes it possible
to define a maximal reconciliation algorithm.

Let therefore the value representing directories be $\vald$,
and let us define subsets of $\setv$ for each type of value:
let $\setd$ be $\{\vald\}$, let $\setb$ be $\{\empt\}$, 
and let $\setf$ contain all remaining values that represent
different file contents together with any metadata associated with them:
\[ \setv = \setb \cup \setf \cup \setd. \]

The tree property can then be defined as
\begin{mydef}[Tree property]
\begin{align*}
\forall n\in\setn:{} &\FS(n) \neq \empt \\
&\quad\wedge \parent(n) \neq \topnode \\
&\Rightarrow \FS(\parent(n)) \in \setd. 
\end{align*}
\end{mydef}
This means that as we move down from the root of a tree of nodes,
the types of values we encounter in the filesystem can only change according to the
transition diagram in \cref{fig_transition}, from directories ($\cchard$) to files ($\ccharf$)
to empty nodes ($\ccharb$).

@@

In this paper, $n$, $m$ and $o$ denote nodes in $\setn$.
We use $\valf$ for an arbitrary element in $\setf$, 
and by $\valvx$ we mean a value of type $X$ in $\setvx{X}$.
